---
title: "Meeting: A.B."
subtitle: ""  
author: "Hans Martinez"
date: '`r format(Sys.Date(),"%b %d, %Y")`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#4F2683",
  secondary_color = "#4F2683",
  inverse_header_color = "#FFFFFF",
  base_font_size = "30px",
  title_slide_background_image = "Horizontal_Rev.png",
  title_slide_background_size = "5% no-repeat",
  title_slide_background_position = "bottom",
  extra_css = list(
  ".small" = list("font-size" = "80%"),
  ".full-width" = list(
    display = "flex",
    width   = "100%",
    flex    = "1 1 auto"
    )
  ),
  header_font_google = google_font("Merriweather", "400","400i","700"),
  text_font_google   = google_font("Arimo", "400", "400i"),
  code_font_google   = google_font("IBM Plex Mono")
)
```

## Objective

- Work-in-progress idea
- Feasibility
- Data

---
layout: false
class: inverse center middle

# Immigrant job market (IJB) and productivity losses in Canada

---
layout: true

### IJB + productivity losses

---

* Immigrants are critical for the economic development of Canada by filling in the gaps in the labor market. By 2036, 30% of pop. will be immigrants .small[(Mortency et al., 2017)]

* There is a positive association between the share of immigrants in a firm and the firm's productivity. High-skilled immigrants are associated with higher firm profits .small[(Gu et al., 2020)], and innovation .small[(Blit et al., 2021)].

* However, immigrants suffer unemployment and underemployment .small[(Adamuti-Trache, 2016)].

---

* On the other hand, Firms face higher costs when hiring an immigrant (non-PR, TFW): LMIA ($1000, complex and inconsistent process, long time) .small[(HC, 2016)] 

* Firms believe native candidates are more competitive than foreign; however, SIEs have more education and experience on average .small[(Chhinzer and Oh, 2021)].

* Likewise, foreign work experience is discounted, while Canadian experience is more valued .small[(Chhinzer and Oh, 2021)]. 

* Furthermore, employers do not actively recruit immigrants (PR, SIE), cannot assess foreign credentials, and do not evaluate immigrants' competencies .small[(Chhinzer and Oh, 2021)]. 

* Firm side remains understudied. 

---

* **Question**: Frictions on both sides of the Canadian job market for immigrants (TFW and SIE) negatively affect firms' productivity. How significant are these losses? Can we measure them?

* **Data**: StatsCan employer-employee + IRCC linked databases

* **Approach**: 
  1. Search model to calibrate frictions and recover counterfactual matches.
  2. 2SGMM procedure for Production function (GNR) + productivity monotonic in immigrant skills.

---
layout: false
class: center, middle

# Thanks!

